---
title: "My Data Viz Work"
author: "Halina Do-Linh"
date: "8/18/2021"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
# General use 
library(tidyverse)
library(here)
library(janitor)

# Specific for data viz
library(patchwork)
library(ggrepel)
library(gghighlight)
library(paletteer)
library(ggExtra)
library(ggbeeswarm)

# Data for interactive session today
library(gapminder)

# Use for working with spatial data in R
library(sf)
```

## Read in lizard data
```{r}
lizards <- read_csv(here::here("data_tidy", "lizards.csv"))

# check the starting point where you are use here() in the console
```

### Making a terrible histogram of lizard weights
```{r}
ggplot(data = lizards, aes(x = weight)) +
  geom_histogram(fill = "orange", 
                 color = "navy blue",
                 size = 1.2,
                 linetype = "dashed")

# top line of ggplot is the base line of expectations, but you could put (data = lizards, aes(x = weight)) into geom_histogram. This allows you to pull multiple data sets to plot (not a common occurrence, but an option).
# color refers to points and lines
# fill refers to the actual fill color for shapes or polygons
# size for polygons refers to the width of the border
# linetype changes the line type
# these are outside of aes() because they are NOT dependent on a variable
```

## Line Plot: total_length on x-axis and weight on y-axis
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight, color = weight)) +
  geom_point(alpha = 0.5, shape = 4)

# shape has a range of 0-25 where each number is a different shape 
# alpha has to do with opacity  
```

## Bubble Plot
Where the color of the points change based on common_name and size of point varies based on total_length.

DISCLAIMER: Bad idea!
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name, size = total_length),
             shape = 22,
             fill = "gray",
             size = 1) +
  theme_void()

# again add color and other styles in aes if the style is based on a variable.
```

## Faceting

Facet wrap creates a grid with two variables (typically categorical). 
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = common_name),
             size = .5) +
  theme_light() +
  facet_wrap(~common_name, ncol = 10)

# ~ = "by" so facet_wrap by common_name
# see differences btwn facet wraps using ncol()
```

Create a facet_grid instead
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) + 
  geom_point(aes(color = common_name)) +
  theme_light() +
  facet_grid(sex~tail)

# facet_grid creates a grid for you based on variables you tell it
 
```

## More examples

Plot lizard counts by common_name. Find total lizard counts by common_name.
```{r}
lizard_counts <- lizards %>% 
  group_by(common_name) %>% 
  summarize(count = n())

lizard_counts

# shortcut to do the same thing: 
# lizard_counts <- lizards %>% dyplr::count(common_name)
# lizard_counts <- lizards %>% count(common_name, tail)
# use the first way when you're trying to find multiple summaries like mean, sd, and count.
# tally() uses summarize and n() but NOT group_by()
```

```{r}
ggplot(data = lizard_counts, aes(y = fct_reorder(common_name, count), x = count)) +
  geom_col(aes(fill = common_name)) +
  labs(x = "Lizard Counts", y = "Lizard Common Name")

# wrote the code this way because having common_name on the y axis makes the name more readable. Switch it and see how it overlaps and becomes unreadable on the x-axis.
# you can also use coord_flip() to switch the x and y variables
# fct_reorder() is built to create a factor and have it reorder based on another variable. Need to tell it the variable. 
# forcats is part of the tidyverse
```

Try converting common name to an ordered factor outside of ggplot
```{r}
lizard_cm_chr <- lizard_counts %>% 
  mutate(common_name = as.factor(common_name))

# can also be written as
# mutate(common_name = fct_reorder(common_name, count))
# then use levels(lizard_cm_chr$common_name)
# you want it be fct_reorder so that it's in the order you want and then check the order using levels

class(lizard_cm_chr$common_name)
```

## Axis scale breaks, limits & labels

Create a scatter plot of total_length (x) by weight(y)
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  scale_x_continuous(breaks = c(0, 50, 500),
                     limits = c(0, 500),
                     expand = c(0, 0)) +
  scale_y_continuous(expand = c(0, 0))

# breaks refers to tick mark breaks
```

```{r}
lizard_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  count(date)

# count(date) and also be written using group_by() and summarize(count = n())
# class(lizard_counts$date)
```

Now let's make the plot
```{r}
ggplot(data = lizard_counts, aes(x = date, y = n))+
  geom_line() +
  scale_x_date(date_breaks = "3 years",
               date_labels = "%y")

# needed to change the date column values from characters to date so that the graph would recognize the date value instead recognizing it as a character
# know what axis you're working on!
```

```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point(aes(color = weight)) +
  scale_color_stepsn(colors = c("green", "red", "purple", "blue"),
                     breaks = c(0, 20, 40, 50, 80))
  
#  scale_color_gradientn(colors = c("orange", "red", "purple", "black"))

# check out other scale_color_gradient functions
```

### Update a color scheme using a palette in paletteer

Make a horizontal boxplot with common_name on the y-axis, total_length on the x-axis, with color changing based on common_name
```{r}
lizards_fct <- lizards %>% 
  mutate(common_name = fct_reorder(common_name, total_length, .fun = median))

# take common_name, turn it into a factor, and then reorder factor values by median of total_length

ggplot(data = lizards_fct, aes(y = common_name, x = total_length)) +
  geom_boxplot(aes(fill = common_name), show.legend = FALSE) +
  labs(y = "Lizard Common Name", x = "Length") +
  scale_fill_paletteer_d(palette = "futurevisions::earth")
  
```

### Yet Another Graph
```{r}
ggplot(data = lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  theme(
    panel.grid = element_blank(),
    plot.background = element_rect("cyan"),
    axis.text.x = element_text(color = "orange"),
    axis.title = element_text(color = "yellow", size = 12)
  )

# the most specific level trumps all other commands in theme. So panel.grid.major.x trumps panel.grid.major. Order does not matter.
```


## Day 9 `ggplot` continued

Read in lizards data again and ran attached packages (library).

## ggrepel

Package for adding labels on a plot.

Make a subset from lizards, called ww_lizards, that only contains observations for "western whiptail" lizards from the site "sand".
```{r}
ww_lizards <- lizards %>% 
  filter(common_name == "western whiptail" & site == "sand") 

```

Making a scatter plot of weight by total_length

```{r}
ggplot(data = ww_lizards, aes(x = total_length, y = weight)) +
  geom_point() +
  geom_text_repel(aes(label = toe_num), size = 3, max.overlaps = 20)

# want to add a label for toe_num
```
## Gapminder Data

Create a scatter plot of gdpPercap (x) and lifeExp (y), labeled by country name. Filter for only Europe.

```{r}
gapminder_euro <- gapminder %>% 
  filter(continent == "Europe", year == 2002) 

ggplot(data = gapminder_euro, aes(x =lifeExp, y = gdpPercap)) +
  geom_point() +
  geom_text_repel(aes(label = country), size = 3)

```

## gghighlight

```{r}
p <- ggplot(lizards, aes(total_length, weight)) +
  geom_point()

p + gghighlight(toe_num == 250, label_key = toe_num)
```

Create a line plot of total_length by weight

```{r}
q <- ggplot(lizards, aes(total_length, weight)) +
  geom_line(aes(color = common_name)) +
  gghighlight(max(weight) > 30)

q
```

## Patchwork for compound figures

patchwork is about using multiple files in one graph

```{r}
 q / (p | q) &
  theme_bw() &
  theme(
    panel.grid.minor = element_blank()
  )

# apply pemdas 
# use the & to add styles to all plots
```

## A Few New Graph Types

### Maringal plots

```{r}
whiptails <- lizards %>% 
  filter(common_name == "western whiptail") %>% 
  drop_na(weight, total_length)

ggplot(data = whiptails, aes(total_length, weight)) +
  geom_point() +
  geom_rug()
```

Marginal Plot with box plots
```{r}
my_plot <- ggplot(data = whiptails, aes(total_length, weight)) +
  geom_point(aes(color = sex), size = 2) +
  scale_color_manual(values = c("cyan4", "black", "goldenrod"),
                     name = "Sex:",
                     labels = c("female", "juvenile", "male")) +
  theme_minimal() +
  theme(
    legend.position = "bottom")

ggMarginal(my_plot, type = "densigram", groupColour = TRUE)

# ggMarginal only uses the types: density, histogram, boxplot, violin, and densigram
```
## Beeswarm Plot

```{r}
ggplot(data = whiptails, aes(sex, weight)) +
  geom_beeswarm() +
  geom_boxplot(fill = NA)
```

### Heatmap with geom_tile

Make a new df called lizard_counts starting from lizards with date converted to class date then count by year and common name the number of lizards observed.
```{r}
lizard_counts <- lizards %>% 
  mutate(date = lubridate::mdy(date)) %>% 
  mutate(date = lubridate::year(date)) %>% 
  count(common_name, date)
#  group_by(year, common_name)
#  summarize(count = c(year, common_name)) this is wrong need to fix it

# for pulling out year, make a new column because you don't want to completely alter or lose your raw data
```

Make the heatmap

```{r}
ggplot(data = lizard_counts, aes(x = date, y = common_name)) +
  geom_tile(aes(fill = n)) +
  geom_text(aes(label = n), color = "white", size = 3) +
  scale_fill_viridis_c()
#  scale_fill_gradientn(colors = c("navy", "red", "orange"))

#scale_fill_viridis is color blind friendly
# this could be a good plot to investigate trends in your data when you are first exploring your data
# color changes line and points
# fill changes the whole fill color
```

### Make a map!

```{r}
jornada_vegetation <- read_sf(here("data_raw", "spatial_vegetation", "doc.kml")) %>% 
  select(Name) %>% 
  clean_names()

ggplot(data = jornada_vegetation) +
  geom_sf(aes(fill = name), color = NA, size = 0.2) +
  scale_fill_paletteer_d(palette = "ggthemes::manyeys") +
  labs(x = "Longitude",
       y = "Latitude",
       fill = "Dominant vegetation:")

# thin while border around polygons works really well or NA bc that's typically how we see maps
# palettes are built specifically for continuous and discrete variables 

```

